Supporting information: Tail associations and their impact on extinction risk

### 2019-11-06 SuppMat.Rmd
small edits on final read through

<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. \ref{MT-fig_intro_cop}B) was 
generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and @Ghosh_copula. A
2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{MT-fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was `r srho_common`, the same as that of the
extreme left-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula. The survival Clayton copula is a rotation
of the Clayton copula by 180 degrees.

@@

$0.875$. Let $u_i$, $v_i$ be the ranks of a sample of noise, $i=1,\ldots,N$.

@@

associated noise, the right-tail case being similar. Then 

@@

\rho=\frac{\sum_{i=1}^N (u_i - \mean(u_i))(v_i-\mean(v_i))}{(N-1)\sqrt{\var(u_i)\var(v_i)}}.

@@

\noindent However, it is straightforward to see that $\mean(u_i) = \mean(v_i)=(N+1)/2$,

@@

that $\var(u_i)=\var(v_i)=\frac{\sum_{i=1}^N (i-\mu)^2}{N-1}$, since 
the set consisting of the $u_i$ is the same as the set consisting of the 
$v_i$, and is also the same as the set consisting of the numbers $1,\ldots,N$. 

@@

\rho=\frac{\sum_{i=1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2}.

@@

\noindent We can reindex the $u_i$ and $v_i$ so that $u_i=v_i=i$ for $i=1,\ldots,N/2$.
This uses the fact that we are working with the left-tail associated noise.

@@

\rho &= \frac{\sum_{i=1}^{N/2} (i-\mu)^2 +\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2} \\
&= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2}.\label{eq:rho1}

@@

\noindent We define $\mu_2 = \mean_{i=N/2+1}^N(u_i)=\mean_{i=N/2+1}^N(i)$. It is straightforward

@@

\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu) &= \sum_{i=N/2+1}^N (u_i-\mu_2+\mu_2-\mu)(v_i-\mu_2+\mu_2-\mu) \\
&= \sum_{i=N/2+1}^N \left( (u_i-\mu_2)(v_i-\mu_2)+(u_i-\mu_2)(\mu_2-\mu)
  +(v_i-\mu_2)(\mu_2-\mu)+(\mu_2-\mu)^2 \right) \\
&= \sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2) + \frac{N}{2}(\mu_2-\mu)^2 \\
&= \sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2) + \frac{N^3}{32}.

@@

\rho &= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (u_i-\mu)(v_i-\mu)}{\sum_{i=1}^N (i-\mu)^2} \\
&= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2) + \frac{N^3}{32}}{\frac{N(N+1)(N-1)}{12}} \\
&= \frac{1}{2}+\frac{12\sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2)}{N(N+1)(N-1)}+

@@

$\cov_{i=N/2+1}^N (u_i,v_i)$. To see this, note that

@@

\cov_{i=N/2+1}^N (u_i,v_i) &= \frac{\sum_{i=N/2+1}^N(u_i-\mu_2)(v_i-\mu_2)}{(N/2-1)

@@

&= \frac{\sum_{i=N/2+1}^N(u_i-\mu_2)(v_i-\mu_2)}{\sum_{i=N/2+1}^N (i-\mu_2)^2}.\label{eq:rho5}

@@

\cov_{i=N/2+1}^N (u_i,v_i) &= \frac{96\sum_{i=N/2+1}^N(u_i-\mu_2)(v_i-\mu_2)}{N(N+2)(N-2)}.

@@

\frac{12\sum_{i=N/2+1}^N (u_i-\mu_2)(v_i-\mu_2)}{N(N+1)(N-1)} = \cov_{i=N/2+1}^N (u_i,v_i) \left( \frac{N(N+2)(N-2)}{8N(N+1)(N-1)} \right).

>>>>

<!--moderate tail association noise-->
Bivarite environmental noise with moderate left-tail association (as in Fig. \ref{MT-fig_intro_cop}B) was 
generated using a 2-dimensional Clayton copula. The statistical
theory of copulas has been introduced systematically in several texts 
(e.g., @nelsen2006_copula; @joe2014_dependence), and was briefly 
introduced by @Genest2007. Short, conceptual introductions aimed at ecological
audiences were provided by sections within the papers of @Anderson2018 and @Ghosh_copula. A
2-dimensional Clayton copula is equivalent to a particular bivariate 
probability density function with marginal distributions which are 
uniform on the interval $(0,1)$. The points on Fig. 
\ref{MT-fig_intro_cop}B were independent draws from the associated random variable.
The functions `claytonCopula` and `rCopula` from the `copula` package for the 
R programming language were used. The parameter of the Clayton family was
chosen so that the Spearman correlation was `r srho_common`, the same as that of the
extreme left- and right-tail associated noise. This was accomplished with the `iRho`
function of the `copula` package. Noise with moderate right-tail
association was generated in a similar way but using the 2-dimensional 
survival Clayton copula. The survival Clayton copula is a rotation
of the Clayton copula by 180 degrees.

@@

$0.875$. Let $x_i$, $y_i$ be the ranks of a sample of noise, $i=1,\ldots,N$.

@@

associated noise, the right-tail case being similar. Then we have

@@

\rho=\frac{\sum_{i=1}^N (x_i - \mean(x_i))(y_i-\mean(y_i))}{(N-1)\sqrt{\var(x_i)\var(y_i)}}.

@@

\noindent However, it is straightforward to see that $\mean(x_i) = \mean(y_i)=(N+1)/2$,

@@

that $\var(x_i)=\var(y_i)=\frac{\sum_{i=1}^N (i-\mu)^2}{N-1}$, since 
the set consisting of the $x_i$ is the same as the set consisting of the 
$y_i$, and is also the same as the set consisting of the numbers $1,\ldots,N$. 

@@

\rho=\frac{\sum_{i=1}^N (x_i-\mu)(y_i-\mu)}{\sum_{i=1}^N (i-\mu)^2}.

@@

\noindent We can reindex the $x_i$ and $y_i$ so that $x_i=y_i=i$ for $i=1,\ldots,N/2$.
This uses the fact that we are working with the extreme left-tail associated noise.

@@

\rho &= \frac{\sum_{i=1}^{N/2} (i-\mu)^2 +\sum_{i=N/2+1}^N (x_i-\mu)(y_i-\mu)}{\sum_{i=1}^N (i-\mu)^2} \\
&= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (x_i-\mu)(y_i-\mu)}{\sum_{i=1}^N (i-\mu)^2}.\label{eq:rho1}

@@

\noindent We define $\mu_2 = \mean_{i=N/2+1}^N(x_i)=\mean_{i=N/2+1}^N(i)$. It is straightforward

@@

\sum_{i=N/2+1}^N (x_i-\mu)(y_i-\mu) &= \sum_{i=N/2+1}^N (x_i-\mu_2+\mu_2-\mu)(y_i-\mu_2+\mu_2-\mu) \\
&= \sum_{i=N/2+1}^N \left( (x_i-\mu_2)(y_i-\mu_2)+(x_i-\mu_2)(\mu_2-\mu)
  +(y_i-\mu_2)(\mu_2-\mu)+(\mu_2-\mu)^2 \right) \\
&= \sum_{i=N/2+1}^N (x_i-\mu_2)(y_i-\mu_2) + \frac{N}{2}(\mu_2-\mu)^2 \\
&= \sum_{i=N/2+1}^N (x_i-\mu_2)(y_i-\mu_2) + \frac{N^3}{32}.

@@

\rho &= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (x_i-\mu)(y_i-\mu)}{\sum_{i=1}^N (i-\mu)^2} \\
&= \frac{1}{2}+\frac{\sum_{i=N/2+1}^N (x_i-\mu_2)(y_i-\mu_2) + \frac{N^3}{32}}{\frac{N(N+1)(N-1)}{12}} \\
&= \frac{1}{2}+\frac{12\sum_{i=N/2+1}^N (x_i-\mu_2)(y_i-\mu_2)}{N(N+1)(N-1)}+

@@

$\cov_{i=N/2+1}^N (x_i,y_i)$. To see this, note that

@@

\cov_{i=N/2+1}^N (x_i,y_i) &= \frac{\sum_{i=N/2+1}^N(x_i-\mu_2)(y_i-\mu_2)}{(N/2-1)

@@

&= \frac{\sum_{i=N/2+1}^N(x_i-\mu_2)(y_i-\mu_2)}{\sum_{i=N/2+1}^N (i-\mu_2)^2}.\label{eq:rho5}

@@

\cov_{i=N/2+1}^N (x_i,y_i) &= \frac{96\sum_{i=N/2+1}^N(x_i-\mu_2)(y_i-\mu_2)}{N(N+2)(N-2)}.

@@

\frac{12\sum_{i=N/2+1}^N (x_i-\mu_2)(y_i-\mu_2)}{N(N+1)(N-1)} = \cov_{i=N/2+1}^N (x_i,y_i) \left( \frac{N(N+2)(N-2)}{8N(N+1)(N-1)} \right).
